---
title: "Reproducible Research / NOOA - Severe Weather Events"
author: "Myrba"
date: "26/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Assignment: Reproducible Research / Week 4 / Course Project 2**

#### **Introduction**

- Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. 
- Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

#### **Synopsis**
- This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. 
- This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

#### **Main Objective**
- The main objective of this research is to explore the ** NOAA Storm Database ** and find answers with some logic that allow us to understand doubts or questions about ** severe weather events **.


### **Data Processing**
#### **Setting up the environment for R**
##### **<span style="color:blue">Installing Libraries</span>**
```{r echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(R.utils)
library(rmarkdown)
library(knitr)
library(plyr)
library(dplyr)
library(ggplot2)
library(lattice)
library(data.table)
library(grid)
library(gridExtra)
```
##### **<span style="color:blue">Setting up Local Time and Language</span>**
```{r echo=TRUE, results='hide', warning=FALSE, message=FALSE}
Sys.setlocale("LC_TIME", "English")
```

#### Loading and preprocessing the data
##### **<span style="color:blue">Read the base bz2 file "stormData.csv.bz2" that contains the data</span>**
```{r, echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}

dataInput_file <- "stormData.csv.bz2"
if (!file.exists(dataInput_file)) {
  data_URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(data_URL, destfile = dataInput_file)

}
if(!file.exists("stormdata.csv"))
{
  bunzip2("stormData.csv.bz2","stormdata.csv",remove=F)
}
#stormData <- read.csv("stormdata.csv", header=TRUE,sep=",")
```
##### **<span style="color:blue">Activating Cache for storing stormData with Results</span>**
```{r readin, cache=TRUE, results = 'markup', warning = TRUE, message = TRUE}
stormData <- read.csv("stormdata.csv", header=TRUE,sep=",")

glimpse(stormData)
```


```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}
head(stormData)
names(stormData)
```

### **<span style="color:green">Results</span>**


#### **Question 1 :** 
- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

#### **Answer :**
- To answer this question, it is clear that the main focus is on discovering the variables FATALITIES and INJURIES.
- It is from these variables that the information to be analyzed and subsequently plotted will be constructed.
  
##### **<span style="color:blue">Retrieving and Listing Top 10 EVTYPE "Injuries" Data Elements for this Analysis</span>**
```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}
# Reviewing events that cause the most Injuries by Weather Event.

## The Procedure is aggregate the top 10 Injuries by the event type and sort the output in descending order


total_injuries <- aggregate(INJURIES~EVTYPE, stormData, sum)
total_injuries <- arrange(total_injuries, desc(INJURIES))
total10_injuries <- total_injuries[1:10, ]
total10_injuries
```

##### **<span style="color:blue">Retrieving and Listing Top 10 EVTYPE "Fatalities" Data Elements for this Analysis</span>**
```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}
# Reviewing events that cause the most fatalities by Weather Event

## The Procedure is aggregate the top 10 fatalities by the event type and sort the output in descending order

total_fatalities <- aggregate(FATALITIES~EVTYPE,stormData, sum)
total_fatalities <- arrange(total_fatalities, desc(FATALITIES))
total10_fatalities <- total_fatalities[1:10, ]
total10_fatalities
```


##### **<span style="color:blue">Plots of Fatalities and Injuries by Events</span>**
```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}
#Show plot of fatalities by Event Type
fatalities_data <- ggplot(total10_fatalities, aes(x = EVTYPE, y = FATALITIES)) + 
      geom_bar(stat = "identity", fill = "darkmagenta", width = NULL) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      xlab("Event Type") + ylab("Fatalities") 
#plot(fatalities_data)

#Show plot of injuries by Event Type
injuries_data <- ggplot(total10_injuries, aes(x = EVTYPE, y = INJURIES)) + 
      geom_bar(stat = "identity", fill = "darkred", width = NULL) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      xlab("Event Type") + ylab("Injuries") 

#plot(injuries_data)

grid.arrange(fatalities_data, injuries_data, ncol=2, nrow=1, bottom = "Figure 1: Comparative Public Health Impact",
     top = textGrob("Public Health Impact - Fatalities & Injuries from top 10 Weather Events", gp=gpar(fontsize=14,font=3)))
```

- As can be seen in the previous figure, by far the event responsible for causing more damage in <span style="color:blue">fatalities</span> and <span style="color:blue">injuries</span> is the one corresponding to **<span style="color:red">Tornado</span>**.  

```{css echo=FALSE}
.custom-inline {
  color: red;
  font-weight: 700
}
```




#### **Question 2 : **
- Across the United States, which types of events have the greatest economic consequences?

#### **Answer :**
- For this question, the analysis is based on the variables Crop (CROPDMG) and Property (PROPDMG) damage.
- These variables are precisely the ones that provide us with information on the damage regarding Crop and Property because of how it harms people, industries and therefore the United States.

##### **<span style="color:blue">Data Explorations and Findings</span>**
- After reviewing the column names, the property damage(PROPDMG) and crop damage(CROPDMG) columns both have another related column titled 'exponents' (i.e - PROPDMGEXP and CROPDMGEXP respectively). 
- As a result, let's convert the exponent columns into numeric data for the calculation of total property and crop damages encountered.

##### **<span style="color:blue">Definitions and Computing</span>**
- Property damage exponents for each level listed out & assigned those values for the property exponent data. 
- Invalid data was excluded by assigning the value as '0'. 
- Then, the property damage value was calculated by multiplying the property damage and property exponent value.


```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}

unique(stormData$PROPDMGEXP)
```

```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}

# Assigning values for the property exponent stormData 
stormData$PROPEXP[stormData$PROPDMGEXP == "K"] <- 1000
stormData$PROPEXP[stormData$PROPDMGEXP == "M"] <- 1e+06
stormData$PROPEXP[stormData$PROPDMGEXP == ""] <- 1
stormData$PROPEXP[stormData$PROPDMGEXP == "B"] <- 1e+09
stormData$PROPEXP[stormData$PROPDMGEXP == "m"] <- 1e+06
stormData$PROPEXP[stormData$PROPDMGEXP == "0"] <- 1
stormData$PROPEXP[stormData$PROPDMGEXP == "5"] <- 1e+05
stormData$PROPEXP[stormData$PROPDMGEXP == "6"] <- 1e+06
stormData$PROPEXP[stormData$PROPDMGEXP == "4"] <- 10000
stormData$PROPEXP[stormData$PROPDMGEXP == "2"] <- 100
stormData$PROPEXP[stormData$PROPDMGEXP == "3"] <- 1000
stormData$PROPEXP[stormData$PROPDMGEXP == "h"] <- 100
stormData$PROPEXP[stormData$PROPDMGEXP == "7"] <- 1e+07
stormData$PROPEXP[stormData$PROPDMGEXP == "H"] <- 100
stormData$PROPEXP[stormData$PROPDMGEXP == "1"] <- 10
stormData$PROPEXP[stormData$PROPDMGEXP == "8"] <- 1e+08

# Assigning '0' to invalid exponent stormData
stormData$PROPEXP[stormData$PROPDMGEXP == "+"] <- 0
stormData$PROPEXP[stormData$PROPDMGEXP == "-"] <- 0
stormData$PROPEXP[stormData$PROPDMGEXP == "?"] <- 0

# Calculating the property damage value
stormData$PROPDMGVAL <- stormData$PROPDMG * stormData$PROPEXP
```

```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}

unique(stormData$CROPDMGEXP)
```

```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}

# Assigning values for the crop exponent stormData 
stormData$CROPEXP[stormData$CROPDMGEXP == "M"] <- 1e+06
stormData$CROPEXP[stormData$CROPDMGEXP == "K"] <- 1000
stormData$CROPEXP[stormData$CROPDMGEXP == "m"] <- 1e+06
stormData$CROPEXP[stormData$CROPDMGEXP == "B"] <- 1e+09
stormData$CROPEXP[stormData$CROPDMGEXP == "0"] <- 1
stormData$CROPEXP[stormData$CROPDMGEXP == "k"] <- 1000
stormData$CROPEXP[stormData$CROPDMGEXP == "2"] <- 100
stormData$CROPEXP[stormData$CROPDMGEXP == ""] <- 1

# Assigning '0' to invalid exponent stormData
stormData$CROPEXP[stormData$CROPDMGEXP == "?"] <- 0

# calculating the crop damage 
stormData$CROPDMGVAL <- stormData$CROPDMG * stormData$CROPEXP
```

```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}

# Property Damage Summary

# Procedure = aggregate the property damage by the event type and sort the output it in descending order

prop <- aggregate(PROPDMGVAL~EVTYPE,stormData, sum)
prop <- arrange(prop, desc(PROPDMGVAL))
prop <- prop[1:10, ]
prop
```

```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}
# Crop Damage Summary

# Procedure = aggregate the crop damage by the event type and sort the output it in descending order

crop <- aggregate(CROPDMGVAL~EVTYPE,stormData, sum)
crop <- arrange(crop, desc(CROPDMGVAL))
crop <- crop[1:10, ]
crop
```


##### **<span style="color:blue">Plots of Damages</span>**
```{r echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}
#Show plot of Property Damages by Event Type
prop_data <- ggplot(prop, aes(x = EVTYPE, y = PROPDMGVAL)) + 
      geom_bar(stat = "identity", fill = "darkmagenta", width = NULL) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      xlab("Event Type") + ylab("Property Damages ($)") 
#plot(fatalities_data)

#Show plot of Crop Damages by Event Type
crop_data <- ggplot(crop, aes(x = EVTYPE, y = CROPDMGVAL)) + 
      geom_bar(stat = "identity", fill = "darkred", width = NULL) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      xlab("Event Type") + ylab("Crop Damages ($)") 

#plot(injuries_data)

grid.arrange(prop_data, crop_data, ncol=2, nrow=1, bottom = "Figure 2: Comparative Damages betweeN Properties and Crops",
     top = textGrob("Public Damage in $ - Properties and Crops from several Weather Events", gp=gpar(fontsize=14,font=3)))
```

#### **Summary of Conclusions: **
- Flooding has resulted in more than double the cost of property/crop damage compared to hurricanes/typhoons. However, the hurricane/typhoon event has only been recorded 88 times, so in terms of cost per event it is extremely costly. Tornado events appear highly again, as it did in terms of population health impact. Extreme heat does not appear despite it having a high health impact, however, drought does appear in the top 10, largely driven by crop damage costs.
- In conclusion, in terms of total impact on health, tornado events have the biggest adverse effects based on fatalities and injuries. However, in terms of the impact per event, excessive heat has a greater effect. Flooding has the biggest total impact on the economy, based on property and crop damage costs. However, hurricane/typhoon events have a greater cost per event.
